---
title: "DATA3888_2025 Lab Week (place number here)"
subtitle: "Place a subtitle here if needed"
author: "(Place your name/Unikey/SID)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    code-tools: true
    code-fold: true
    fig_caption: yes
    number_sections: yes
    embed-resources: true
    theme: flatly
    css: 
      - https://use.fontawesome.com/releases/v5.0.6/css/all.css
    toc: true
    toc_depth: 4
    toc_float: true
execute:
  echo: true
  tidy: true
---

# 1 Load the data

```{r}
# Load the expression matrix with symbols
eMat <- readRDS("data/eMat_with_symbols.rds")

# View dimensions and a preview
dim(eMat)
#head(eMat) 
```

```{r}
fdata <- readRDS("data/fData.rds")
pdata_final <- readRDS("data/pData_final.rds")

#head(pdata_final)
```

```{r}
library(dplyr)  
eMat_t <- t(eMat)

eMat_t_df <- as.data.frame(t(eMat))
eMat_t_df$sample_id <- rownames(eMat_t_df)

pdata_final$sample_id <- rownames(pdata_final)

merged_df <- dplyr::inner_join(pdata_final, eMat_t_df, by = "sample_id")
rownames(merged_df) <- merged_df$sample_id

dim(merged_df)
# head(merged_df)
```

### ðŸ” How to Construct the `flare_next` Label

We aim to train a model that **uses gene expression and clinical data from the current visit to predict whether a flare (SLEDAI â‰¥ 5) will occur at the next visit**.

```{r}
merged_df <- merged_df %>%
  mutate(
    visit_count = as.numeric(visit_count),
    sledai = as.numeric(sledai),
    patient_id = as.character(patient_id)
  )

merged_long <- merged_df %>%
  filter(group_type == "SLE") %>%
  arrange(patient_id, visit_count) %>%
  group_by(patient_id) %>%
  mutate(
    sledai_next = lead(sledai),
    flare_next = ifelse(lead(sledai) >= 5, 1, 0),
    visit_next = lead(visit_count)
  ) %>%
  ungroup() %>%
  filter(!is.na(sledai_next))
```

# 2 Gene-based flare prediction (Current Time Point Only)

```{r}
gene_cols <- setdiff(colnames(merged_long), c(
  "sample_id", "illumina_id", "geo_accession", "group_type", "patient_id",
  "visit", "batch", "batch_replicate", "set", "visit_count", "cumulative_time",
  "days_since_diagnosis", "days_since_last_visit", "days_between_diagnosis_and_last_visit",
  "disease state", "gender", "race", "age", "sledai", "sledai_next", "flare_next", "visit_next"
))

numeric_gene_cols <- gene_cols[sapply(merged_long[, gene_cols], is.numeric)]

cor_vals <- apply(merged_long[, numeric_gene_cols], 2, function(x) cor(x, merged_long$flare_next, use = "complete.obs"))

top_genes <- names(sort(abs(cor_vals), decreasing = TRUE))[1:30]

gene_data_top <- merged_long[, c("flare_next", top_genes)]
```

```{r}
library(caret)

merged_long$flare_next <- as.factor(merged_long$flare_next)

set.seed(123)
train_idx <- createDataPartition(merged_long$flare_next, p = 0.8, list = FALSE)
train_data <- merged_long[train_idx, c(top_genes, "flare_next")]
test_data  <- merged_long[-train_idx, c(top_genes, "flare_next")]

rf_model <- train(
  flare_next ~ ., 
  data = train_data,
  method = "rf",
  trControl = trainControl(method = "cv", number = 5),
  importance = TRUE
)

pred <- predict(rf_model, newdata = test_data)
confusionMatrix(pred, test_data$flare_next)
```

# 3 Longitudinal flare prediction (Current + Previous Visit)

### ðŸ” **How I Predicted the Next Flare**

To predict whether a patient would experience a **flare in the next visit** (defined as SLEDAI â‰¥ 5), I constructed a new binary variable called `flare_next`. This variable was created by grouping visits by patient ID and checking if the **SLEDAI score at the subsequent visit** met the flare threshold:

-   If `next_sledai â‰¥ 5`, then `flare_next = 1` (flareï¼‰

-   If `next_sledai < 5`, then `flare_next = 0` (no flare)

-   If no future visit exists, the row was dropped.

ðŸ§¬ **Features Used**

To build a **longitudinal model**, I used gene expression data from both:

-   The **current visit**, and

-   The **previous visit** (as lagged variables, e.g. `prev_GENE1`)

-   These features were selected based on their **Pearson correlation** with `flare_next`. The **top 30 most correlated genes** (from \~21,000) were included in the model.

ðŸ§  **Model Trained**

I trained a **Random Forest classifier** using the `caret` package in R:

5-fold cross-validation was applied.

The target was `flare_next`.

The model was evaluated on a held-out test set.

ðŸ“ˆ **Performance Summary**

-   **Accuracy**: 67.7%

    **Balanced Accuracy**: 67.6%

    **Kappa**: 0.35 (moderate agreement)

    **Sensitivity** (no flare): 70.8%

    **Specificity** (flare): 64.4%

    **P-value vs baseline (NIR)**: 0.00038 â†’ statistically significant improvement

    These results suggest that **longitudinal gene expression data** can moderately predict upcoming flares, offering potential for early intervention.

```{r}
# Create lag features (previous time point) for top genes only
merged_long <- merged_long %>%
  group_by(patient_id) %>%
  mutate(across(all_of(top_genes), list(prev = ~lag(.)), .names = "prev_{.col}")) %>%
  ungroup()

# Use only top genes + their lagged versions
feature_cols <- c(top_genes, paste0("prev_", top_genes))
train_df <- merged_long[, c("flare_next", feature_cols)]

# Ensure only numeric variables
train_df <- train_df[, sapply(train_df, is.numeric)]
train_df$flare_next <- as.factor(merged_long$flare_next)

# Split and train
set.seed(123)
train_idx <- createDataPartition(train_df$flare_next, p = 0.8, list = FALSE)
train_data <- train_df[train_idx, ]
test_data  <- train_df[-train_idx, ]

train_data <- na.omit(train_data)
test_data  <- na.omit(test_data)

# Train using memory-efficient ranger
library(caret)
rf_model_long <- train(
  flare_next ~ ., 
  data = train_data,
  method = "ranger",
  trControl = trainControl(method = "cv", number = 5),
  importance = "impurity"
)

# Predict and evaluate
pred <- predict(rf_model_long, newdata = test_data)
confusionMatrix(pred, test_data$flare_next)
```
